Meta-analysis of composite outcomes in clinical trials

### 2025-01-28 MSc_poster_template.Rmd
Comments addressed

# Methodology
The meta-analysis of the EMERGE outcomes used permutation tests, which generate a null distribution by randomly shuffling the treatment group assignments of participants and their outcome variables. This breaks any true relationship between treatment and outcomes. Statistical significance is assessed by comparing the actual test statistic to the distribution of test statistics from these shuffled datasets. The P-value is calculated as the proportion of the null distribution that is more extreme than the observed test statistic.

@@

# Early Results 
```{r morefigs, out.width='90%', echo=FALSE, fig.cap='Permuted Null Distribution of Composite. Test statistic 28.89, P-value of P<0.001', fig.height=6}
continuous_permutation_function<-function(x){
  permutation_t_value<-numeric(10000)
  i <-1
  repeat{
    thesis.df$Group<-sample(thesis.df$Group, 535, replace = F)
    permutation_result<-t.test(x[thesis.df$Group == "Metformin"], x[thesis.df$Group == "Placebo"])
    permutation_t_value[i]<-permutation_result$statistic
    i<- i +1
    if(i > 10000) break}
  return(permutation_t_value)
}
squared_composite<-((continuous_permutation_function(thesis.df$binary_insulin))^2) +
  ((continuous_permutation_function(thesis.df$gw32))^2) +
  ((continuous_permutation_function(thesis.df$gw38))^2)
percentile_95<-quantile(squared_composite, 0.95)

squared_composite.df<-data.frame(squared_composite)
ggplot(squared_composite.df, aes(x= squared_composite))+
  geom_histogram(bins = 200, fill = "darkslategrey", color = "darkslategrey")+
  scale_x_continuous(limits = c(0, 30)) +
  labs(x = "Squared Composite", y = "Frequency", title = expression(bold("Permuted Null Distribution"))) +
  geom_vline(xintercept = percentile_95, color = "maroon", linetype = "dashed", size = 1)+
  annotate("text", x = percentile_95, y = 250, label = "95th Percentile", color = "maroon", angle = 90, vjust = -0.5, size = 7)+
  theme_minimal()+
  theme(axis.title = element_text( size = 24),
        axis.text = element_text(size = 18),
        plot.title = element_text(size = 28))+
  geom_vline( xintercept = study_squared_composite, color = "gold3", size = 1 )+
  annotate("text", x = study_squared_composite, y = 250, label = "EMERGE Composite", color = "gold3", angle = 90, vjust = -0.5, size = 7)

@@

```{r morefigs2, out.width='95%', echo=FALSE, fig.cap='Power of Permutation Test', fig.height=5}

@@

The next steps in our project is to compare the permutation method with other
approaches to improving the analysis of composite outputs, in particular to the
methods described by Mc Menamin *et al,* [^5].

@@

The code and datasets for this project can be viewed at our GitHub repository here: <https://github.com/Brendan-G-Woods/Thesis>

>>>>

# Methodology
The meta-analysis of the EMERGE outcomes used permutation tests, which generate a null distribution by randomly shuffling the treatment group assignment of participants, while maintaining the outcome variables association with each other. This breaks any true relationship between treatment and outcomes. Statistical significance is assessed by comparing the actual test statistic to the null distribution from these shuffled datasets. The P-value is calculated as the proportion of the null distribution that is more extreme than the observed test statistic. The composite test statistic is obtained by summing the squared T-values comparing glucose at weeks 32 and 38, and Insulin: ***T~32~^2^ + T~38~^2^ + T~I~^2^ ***

@@

# Early Results 
```{r morefigs, out.width='90%', echo=FALSE, fig.cap='Permuted Null Distribution of Composite. Test statistic 28.89, P-value of P<0.001', fig.height=4.5}
continuous_permutation_function<-function(x){
  permutation_t_value<-numeric(10000)
  i <-1
  repeat{
    thesis.df$Group<-sample(thesis.df$Group, 535, replace = F)
    permutation_result<-t.test(x[thesis.df$Group == "Metformin"], x[thesis.df$Group == "Placebo"])
    permutation_t_value[i]<-permutation_result$statistic
    i<- i +1
    if(i > 10000) break}
  return(permutation_t_value)
}
squared_composite<-((continuous_permutation_function(thesis.df$binary_insulin))^2) +
  ((continuous_permutation_function(thesis.df$gw32))^2) +
  ((continuous_permutation_function(thesis.df$gw38))^2)
percentile_95<-quantile(squared_composite, 0.95)

squared_composite.df<-data.frame(squared_composite)
ggplot(squared_composite.df, aes(x= squared_composite))+
  geom_histogram(bins = 200, fill = "salmon", color = "salmon")+
  scale_x_continuous(limits = c(0, 30)) +
  labs(x = "Squared Composite", y = "Frequency", title = expression(bold("Permuted Null Distribution"))) +
  geom_vline(xintercept = percentile_95, color = "maroon", linetype = "dashed", size = 1)+
  annotate("text", x = percentile_95, y = 200, label = "95th Percentile", color = "maroon", angle = 90, vjust = -0.5, size = 7)+
  theme_minimal()+
  theme(axis.title = element_text( size = 24),
        axis.text = element_text(size = 18),
        plot.title = element_text(size = 28))+
  geom_vline( xintercept = study_squared_composite, color = "gold3", size = 1 )+
  annotate("text", x = study_squared_composite, y = 200, label = "EMERGE Composite", color = "gold3", angle = 90, vjust = -0.5, size = 7)

@@

```{r morefigs2, out.width='95%', echo=FALSE, fig.cap='Power of Permutation Test', fig.height=4.5}

@@

The next steps in our project are to:

* Compare the permutation test method with the approach Mc Menamin *et al.*, [^5] used to analyze composite outputs.
* Compare different methods of creating the composite and their influence on power.
* Derive a theoretical distribution incorporating correlation between the variables.

@@

Our code and datasets can be found here: <https://github.com/Brendan-G-Woods/Thesis>
